An enhancer code compendium of early human neural and neural crest development

### 2025-04-18 manuscript.tex
contributions
make text of fig 1 clearer

%TODO:D) The black outline on green text (EMX2/EN1 etc) makes it difficult to read and see the color. consider making it thinner

@@

\begin{itemize}
  \item Conceptualization: S.D.W. \& S.A.
  \item Experiments and sample preparation: S.D.W., V.C., E.S., L.H. \& C.M.
  \item Computational analysis: S.D.W. \& C.M.
  \item Writing – original draft: S.D.W. \& S.A.
  \item Writing – review \& editing:
\end{itemize}

@@

Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq. Using PCR. A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and 
%TODO: Add concentration
PCR program: \SI{95}{\degreeCelsius} for 3 minutes, 15 cycles of: \SI{98}{\degreeCelsius} for 20 seconds,
\SI{58}{\degreeCelsius} for 15 seconds and \SI{72}{\degreeCelsius} for 2.5 minutes, followed by \SI{72}{\degreeCelsius}
for 5 minutes.
Linarized plasmids were purified using QIAquick PCR purification kit (Qiagen, 28104), followed by DpnI digestion for 1
hour at \SI{37}{\degreeCelsius} and gel extraction of 4,5 kb band using QIAquick Gel Extraction Kit (Qiagen, 28704).
Enhancer candiates were cloned into the target vector using an NEBuilder (New England Biolabs, E5520S) reaction with a linarized plasmid to insert ratio of 1:2 (45 minutes at \SI{50}{\degreeCelsius}). Next, \SI{2.5}{\micro\litre} of the rection mixture was transformed into \SI{25}{\micro\litre} Stellar Competent Cells (Takara Bio, 636763). Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice. 
Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
over night.
Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
concentration of \SI{4}{\micro\gram\per\micro\litre}.

@@

\subsubsection*{Footprinting analysis}
TN5 cut site BigWigs were calculated for all organoid topics using the bigwig command from
scatac\_fragment\_tools ()
%TODO: add RRID
using the --cut-sites option.
Next, for each pattern (converted to a PWM) of the FOX cluster of neural progenitors (cluster 6) motif hits were calculated on the top
1,000 regions of all neural progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
Overlapping hits were resolved by taking the hit with the highest $-log_{10}(pvalue)$. 
For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
region originated, was retained.
FOX patterns were aligned using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) to a seed pattern defined as the pattern with the highest information content.
Finally, regions were centered on the pattern hit position and TN5 cute site coverage was calculated using pyBigWig
(v0.3.24; \href{https://rrid.site/data/record/nlx_144509-1/SCR_024807/resolver}{RRID:SCR\_024807}).

>>>>

@@

\begin{itemize}
  \item Conceptualization: S.D.W. \& S.A.
  \item Experiments and sample preparation: S.D.W., V.C., L.H., C.M., R.M.W., \& R.V.P.
  \item Computational analysis: S.D.W. \& C.M.
  \item Writing – original draft: S.D.W., C.M. \& S.A.
  \item Writing – review \& editing: S.D.W., C.M. \& S.A.
\end{itemize}

@@

Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR. A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and 
PCR program: \SI{95}{\degreeCelsius} for 3 minutes, 15 cycles of: \SI{98}{\degreeCelsius} for 20 seconds,
\SI{58}{\degreeCelsius} for 15 seconds and \SI{72}{\degreeCelsius} for 2.5 minutes, followed by \SI{72}{\degreeCelsius}
for 5 minutes was used.
Linarized plasmids were purified using QIAquick PCR purification kit (Qiagen, 28104), followed by DpnI digestion for 1
hour at \SI{37}{\degreeCelsius} and gel extraction of 4,5 kb band using QIAquick Gel Extraction Kit (Qiagen, 28704).
Enhancer candiates were cloned into the target vector using an NEBuilder (New England Biolabs, E5520S) reaction with a linarized plasmid to insert ratio of 1:2 (45 minutes at \SI{50}{\degreeCelsius}). Next, \SI{2.5}{\micro\litre} of the rection mixture was transformed into \SI{25}{\micro\litre} Stellar Competent Cells (Takara Bio, 636763). Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice. 
Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
over night.
Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
concentration of \SI{4}{\micro\gram\per\micro\litre}.

@@

\subsubsection*{Footprinting analysis}
TN5 cut site BigWigs were calculated for all organoid topics using the bigwig command from
scatac\_fragment\_tools (v0.1.3; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026713/resolver?}{RRID:SCR\_026713})
using the --cut-sites option.
Next, for each pattern (converted to a PWM) of the FOX cluster of neural progenitors (cluster 6) motif hits were calculated on the top
1,000 regions of all neural progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
Overlapping hits were resolved by taking the hit with the highest $-log_{10}(pvalue)$. 
For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
region originated, was retained.
FOX patterns were aligned using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) to a seed pattern defined as the pattern with the highest information content.
Finally, regions were centered on the pattern hit position and TN5 cute site coverage was calculated using pyBigWig
(v0.3.24; \href{https://rrid.site/data/record/nlx_144509-1/SCR_024807/resolver}{RRID:SCR\_024807}).
